---
title: "CDISC/ADaM Survival Analysis Report"
author: "Vladislav Fedorov"
date: "2026-07-09"
output:
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

```{r load-data}
library(haven)
library(dplyr)
library(tidyr)
library(survival)
library(ggpubr)
library(survminer)

adsl <- read_xpt("data/adsl.xpt") |>
  mutate(ARM = factor(ARM, levels = c("Placebo", "Xanomeline Low Dose", "Xanomeline High Dose")))

adae <- read_xpt("data/adae.xpt")

adtte <- read_xpt("data/adtte.xpt") |>
  mutate(event = 1 - CNSR)

adtte_with_arm <- adtte |>
  left_join(adsl |> select(USUBJID, ARM), by = "USUBJID")
```

# 1. Baseline Demographics

```{r table1, echo=TRUE}
table1 <- adsl |>
  group_by(ARM) |>
  summarise(
    n = n(),
    mean_age = mean(AGE),
    sd_age = sd(AGE),
    pct_female = mean(SEX == "F") * 100,
    pct_white = mean(RACE == "WHITE") * 100
  )

table1
```

For this Alzheimer treatment study, we are able to observe demographics across three treatment groups - Placebo, Low, and High doses of Xanomeline. The treatment groups are balanced in size, allowing for a more even analysis, with the age being close in mean across them as well. It is worth noting that percentage of female patients present in the study - ranging from 61.6% in Placebo to 47.6% in High Dose - does differ, potentially due to randomization noise and not a real imbalance. While the initial models in the analysis are univariate, knowing the demographics beforehand will aid in studying age-adjusted models in the later stages of this analysis.  

# 2. Adverse Event Summary

```{r ae-prep}
adae_flag <- adae |>
  distinct(USUBJID) |>
  mutate(has_ae = TRUE)

joined_ad <- adsl |>
  left_join(adae_flag, by = "USUBJID") |>
  mutate(has_ae = replace_na(has_ae, FALSE))
```

```{r ae-summary, echo=TRUE}
ae_summary <- joined_ad |>
  group_by(ARM) |>
  summarise(
    n = n(),
    ae_pct = mean(has_ae) * 100
  )

ae_summary
```

```{r top-aes, echo=TRUE}
top_aes <- adae |>
  left_join(adsl |> select(USUBJID, ARM), by = "USUBJID") |>
  distinct(USUBJID, ARM, AEDECOD) |>
  count(AEDECOD, ARM) |>
  pivot_wider(names_from = ARM, values_from = n, values_fill = 0) |>
  mutate(Total = rowSums(across(c(2:4)))) |>
  arrange(desc(Total)) |>
  slice_head(n = 10)

top_aes
```